Supporting information: Tail associations and their impact on extinction risk

### 2019-11-08 MainText.Rmd
adding new csl
finished incorporating Lawrence comments

---
title: "Tail associations and their impact on extinction risk"
author: "Shyamolina Ghosh, Daniel C. Reuman"
date: ""
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. 
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables. These ideas are revisited in the Discussion.

@@

Because the deterministic, one-patch versions of our models showed undercompensatory
dynamics for $r<r_c$ (and therefore for $r$ on the left end of the range $(r_{min},r_{bf})$),
and showed overcompensatory dynamics for $r>r_c$, Figs \ref{fig_extrisk_allmodel_localdisp}
and \ref{SM-fig_extrisk_allmodel_globaldisp} suggest the hypothesis that the
opposite effects of tail association on extinction risk correspond to the domains
in which models exhibit under- versus overcompensatory dynamics. This was 
approximately, but not exactly, true. Careful inspection of Figs 
\ref{SM-fig_risk_local_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} and Figs \ref{SM-fig_risk_global_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6} 
and the $r_c$ values of Table \ref{tab_model_summary} reveals that, for the Velhurst and Austin-Brewer models,
the switch occurs at a value of $r$ less than $r_c$ for these models. For the example
case of $d=0$, plots of extinction risk versus $r$ (Fig. \ref{fig_extrisk_d0}) seem to indicate that,
for the Ricker, Hassell, and Pennycuick models, left-tail association accentuates and
right-tail association mitigates extinction risk for undercompensatory dynamics ($r<r_c$),
whereas the reverse occurs for overcompensatory dynamics ($r>r_c$). Fig. \ref{fig_extrisk_d0}
reveals a similar reversal for the Maynard Smith, Velhurst and Austin-Brewer models; but 
it occurs at a value of $r$ close to but distinguishable from $r_c$.

@@

<!--linking back to ECEs-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and even specific events can now be attributed to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechanisms we explored in this paper. 
If $X_i(t)$ denotes the value of an environmental variable (e.g., a summer maximum 
temperature, or winter minimum temperature, spring rainfall, etc.) 
in location $i$ in year $t$,
@ummenhofer2017extreme provide a diagram of how ECEs may be increasing in frequency 
and intensity, linked to changes in the distribution of $X_i(t)$ such as changes
in the mean, variance, or skewness of this distribution. Although changes in 
the correlation between $X_i(t)$ and $X_j(t)$ are increasingly documented, 
and may also be a consequence of climate change (e.g., @Black2018),
we are unaware of any studies examining the extent to which tail associations
between $X_i(t)$ and $X_j(t)$, and asymmetries of tail association, may be 
changing. However, documented changes in correlation between $X_i(t)$ and $X_j(t)$
may be due partly or wholly to changes in tail associations. If ECEs are becoming
more spatially extensive, as seems likely given that they are becoming
more intense [@ummenhofer2017extreme], then the conditional probability that both of
$X_i(t)$ and $X_j(t)$ are influenced by an ECE, given that one of them is so
influenced, should be increasing. This should contribute to increased association
of the two variables in one of their tails (which tail depends on the type of 
ECE and what the variable $X$ is). A study of changes through time in 
tail associations of weather variables would be straightforward and 
potentially important given the mechanisms we have explored.
<!--***DAN: Shyamolina, another aspect of an exploration you may do on the 
the geography of tail association may be to look at how tail association 
may be changing. This relates to ECEs, so may help better justify the work
to a funder. You had proposed to look at geography of tail association in 
NDVI, and that makes good sense. You could also look at how tail association
is changing in NDVI (the records are about 35 years long, so you could just
about do this). Even better (because data are better), you could get weather
station records (lots of these can be downloaded, over 1000 stations
have been operating for over a century continuously in North America) and 
look at how tail associations are changing in those data.-->

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between the values of an environmental variable in different
locations, $X_i(t)$ and $X_j(t)$, using standard methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used are recommended instead.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even though Pearson correlation
appears to indicate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends not only on association information, but also on the marginals.
For this reason, it is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of 
variables, rather than reflecting true changes in association (these studies
used Pearson correlations).
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, including how such 
associations may be changing.

<!--Tuljapurkar approximation-->
Our results suggest that tail associations and asymmetries of tail associations
may have limited influence on the extinction risk of 
stage-structured, non-spatial models. Although we believe more research is 
certainly still needed
as to whether this conclusion holds more generally than the model we considered
(see below), if it does hold 
generally, it may be possible to obtain a better understanding of 
the result using a well-known approximation
of Tuljapurkar [@tulja1982; @tulja1990] for the long-term stochastic growth rate.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extintion risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations.
We also note that this reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.

<!--***DAN: decided the below was a step too far
Finally, it seems likely that the 
stochastic matrix modelling framework, and Tuljapurkar's approximation,
should apply to the density-independent metapopulation model of 
@Ghosh_copula, which was comprised of Lewontin-Cohen models linked 
by dispersal. Nevertheless asymmetry of tail associations made a 
big difference to extinction risk for that model [@Ghosh_copula].
We have not examined whether Tuljapurkar's approximation is
accurate in that spatially explicit case.

***DAN: Dan to maybe add another para on the fact that I thnk Tulja's approx should also
work for our spatial, non-dd model (check first, Dan), but we see a strong effect here,
so not clear why. And also suggests the possibilty that for some age-structured models,
tail associations will make a bigger difference then they did in our case. Maybe somehow
get to the suggestion that someone needs to revisit Tulja's approx and the analytic
approximation of extinction risk of Lande and Orzack looking at tail associations?

***DAN: In Discussion need to talk about what analytic results are needed
to take the next step of determining when tail depencence may make enough of
a difference to think about. I think our results, together with Tulja's 
approximation, suggest that tail associations probably don't make much difference
for models without a spatial component. But our results illsutaret that it can
make a big difference when there is a spatial component, apparently because of 
the influence of tail associations on rescue effects. Could maybe even outline
analytic approaches for non-dd models? Not sure what would be done with dd
models. The point is you need to take the next step in the Discussion of 
talking about how our results, which admitedly are just for specific models,
could be turned into analytic results or general results. Interestingly, 
Tulja's approximation may also apply to the spatial case, at least when 
all locations can be reached from all other locations via dispersal. But the 
approximation seems poor in that case. Need to think about it more, and maybe
need more research.-->

@@

<!--future ideas-->
In our opinion, additional research would be worthwhile on the 
impact of tail associations on growth rates and 
extinction risk under additional modelling scenarios not considered
in this paper. First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, relating to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be well be associated with,
for instance, $r_2$, whereas lower values of $r_1$ may not be strongly
associated with $r_2$ because reductions/increases in $r_1$ can
then be traded off against increases/reductions in *either* 
$r_2$ or $r_3$. Second, for our metapopulation models,
we assumed habitat patches were equispaced on a transect or
were all equally mutually accessible through dispersal.
Furthermore, associations between environmental variables in different
patches were assumed to be spatially homogeneous. In real metapopulations,
neither of these assumptions holds. Connecting the ideas of tail association that
we have developed in this study with the ideas on the geography of synchrony
of @walteretal2017 seems like a promising future direction.

>>>>

---
title: "Tail associations and their impact on extinction risk"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Daniel C. Reuman"
date: ""
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

\noindent Sheppard: Department of Ecology and Evolutionary Biology and Kansas Biological Survey, University of Kansas, Lawrence, KS, 66045, USA

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. The variance, $\var$,
here and throughout, is the sample variance, with Bessel's correction (denominator $T-1$).
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables. These ideas are revisited in the Discussion.

@@

Because the deterministic, one-patch versions of our models showed undercompensatory
dynamics for $r<r_c$ (and therefore for $r$ on the left end of the range $(r_{min},r_{bf})$),
and showed overcompensatory dynamics for $r>r_c$, Figs \ref{fig_extrisk_allmodel_localdisp}
and \ref{SM-fig_extrisk_allmodel_globaldisp} suggest the hypothesis that the
opposite effects of tail association on extinction risk correspond to the domains
in which models exhibit under- versus overcompensatory dynamics. This was 
approximately, but not exactly, true. Careful inspection of Figs 
\ref{SM-fig_risk_local_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6} and Figs \ref{SM-fig_risk_global_disp_verhulst_r_vary_scl_0.6} - \ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6} 
and the $r_c$ values of Table \ref{tab_model_summary} reveals that, for the Verhulst and Austin-Brewer models,
the switch occurs at a value of $r$ less than $r_c$ for these models. For the example
case of $d=0$, plots of extinction risk versus $r$ (Fig. \ref{fig_extrisk_d0}) seem to indicate that,
for the Ricker, Hassell, and Pennycuick models, left-tail association accentuates and
right-tail association mitigates extinction risk for undercompensatory dynamics ($r<r_c$),
whereas the reverse occurs for overcompensatory dynamics ($r>r_c$). Fig. \ref{fig_extrisk_d0}
reveals a similar reversal for the Maynard Smith, Verhulst and Austin-Brewer models; but 
it occurs at a value of $r$ close to but distinguishable from $r_c$.

@@

<!--linking back to ECEs-->
Extensive evidence now exists that ECEs have become more frequent and more intense
over the last several decades, and substantial progress has recently been made so that
some event categories and even specific events can now be attributed to
human-induced climate change [@ummenhofer2017extreme]. We here provide a fuller 
explanation of how such changes may modify the asymmetry of tail association of 
weather variables measured in different places, and thereby influence metapopulation
extinction risk through the mechanisms we explored in this paper. 
Although changes in 
the correlation between $X_i(t)$ and $X_j(t)$ are increasingly documented, 
and may also be a consequence of climate change (e.g., @Black2018),
we are unaware of any studies examining the extent to which tail associations
between $X_i(t)$ and $X_j(t)$, and asymmetries of tail association, may be 
changing. However, documented changes in correlation between $X_i(t)$ and $X_j(t)$
may be due partly or wholly to changes in tail associations. If ECEs are becoming
more spatially extensive, as seems likely given that they are becoming
more intense [@ummenhofer2017extreme], then the conditional probability that both of
$X_i(t)$ and $X_j(t)$ are influenced by an ECE, given that one of them is so
influenced, should be increasing. This should contribute to increased association
of the two variables in one of their tails (which tail depends on the type of 
ECE and what the variable $X$ is). A study of changes through time in 
tail associations of weather variables would be straightforward and 
potentially important given the mechanisms we have explored.
<!--***DAN: Shyamolina, another aspect of an exploration you may do on the 
the geography of tail association may be to look at how tail association 
may be changing. This relates to ECEs, so may help better justify the work
to a funder. You had proposed to look at geography of tail association in 
NDVI, and that makes good sense. You could also look at how tail association
is changing in NDVI (the records are about 35 years long, so you could just
about do this). Even better (because data are better), you could get weather
station records (lots of these can be downloaded, over 1000 stations
have been operating for over a century continuously in North America) and 
look at how tail associations are changing in those data.-->

<!--copulas and rank methods-->
It is known that the distributions $X_i(t)$ are changing, both in
mean and variance, and possibly in other ways, depending on what
environmental variable $X$ represents.
Studying associations between $X_i(t)$ and $X_j(t)$ using standard 
methods such as Pearson correlation
may be complicated by these changes; rank-based and related 
measures of association such as 
we have used may be more appropriate for some applications.
If $f$ is a strictly monotonic, increasing transformation, then the Spearman
(or Kendall) correlation of $f(X_i(t))$ and $f(X_j(t))$ is the same as the
Spearman (or Kendall) correlation of $X_i(t)$ and $X_j(t)$. However, the Pearson
correlation of $f(X_i(t))$ and $f(X_j(t))$ may differ substantially
from the Pearson correlation of $X_i(t)$ and $X_j(t)$. 
Relatedly, if the marginal distributions of $X_i(t)$ and $X_j(t)$ are
changing through time, it may also cause changes through time in the 
Pearson correlation of these quantities, even while the values
of rank-based correlations such as Spearman and Kendall may be unchanging.
Essentially, in such a case, associations between $X_i(t)$ and $X_j(t)$
would be unchanging through time even though Pearson correlation
appears to indicate a change in association, which is instead a change 
in marginal distributions.
Perhaps unfortunately, the Pearson correlation,
which is the most commonly used measure of association of two variables,
depends not only on association information, but also on the marginals.
For this reason, it is possible that recently observed changes in environmental
(e.g., @Black2018) and biological (e.g., @Koenig2016) synchrony 
are, at least in part, just another feature of 
already-studied changes in distributional characteristics of 
variables, rather than reflecting true changes in association as would be measured
using rank-based methods (these studies used Pearson correlations).
Rank-based methods are linked to a well-developed suite of statistical
approaches related to the *copula* concept of statistics. Copulas and related ideas
can be used to formally separate the information in the joint 
distribution of $X_i$ and $X_j$ into the information that pertains
to the marginal distributions of these two variables, and the information
that pertains solely to their association. 
@Ghosh_copula argued generally for the potential of copulas and related ideas
to improve our understanding of ecology 
(see also @Popovic2018; @Valpine2014; @Anderson2018).
Rank-based methods such as the Spearman and partial Spearman correlations
which we have used are closely related to copula approaches, and @Ghosh_copula,
@Genest2007 and others have recommended copula and rank approaches when the 
goal is to understand associations between variables, including how such 
associations may be changing. We recognize that the Pearson correlation
of $X_i(t)$ and $X_j(t)$, and changes through time in that correlation, 
will still be and appropriate and important quantity to study 
for many applications. For instance, in cases for which $X_i$ has a linear 
or approximately linear influence on some aspect of the ecosystem
at location $i$, Pearson correlations of the ecological impacts at locations 
$i$ and $j$, which may be of fundamental interest, will relate to 
Pearson correlations between $X_i$ and $X_j$. But in many cases the
ecological impact of $X_i$ and $X_j$ will be a nonlinear or unknown monotonic functions of 
the values of $X_i$ and $X_j$. So studying rank-based correlations
may be more informative, ecologically as well as providing the mathematical advantages
listed above.

<!--future ideas start here-->
Our results suggest some new lines of research on the 
impact of tail associations on growth rates and 
extinction risks. First, we considered a model with positive association between 
egg and adult vital rates,
corresponding to similar impacts of an environmental variable on 
eggs and adults. Negative associations between vital rates
can occur as well, and the impacts of tail associations should be considered
in that context as well. Negative associations relate to
life history trade-offs, e.g. between survival and reproduction 
(e.g, @Lande2003, p. 59). Asymmetric tail associations may occur
in such a context if, for instance, there is a three way trade-off
among vital rates $r_1$, $r_2$, and $r_3$. Values of $r_1$ near
an upper limit may be strongly associated with,
for instance, $r_2$, whereas lower values of $r_1$ may not be strongly
associated with $r_2$ because reductions/increases in $r_1$ can
then be traded off against increases/reductions (respectively) in either 
$r_2$ or $r_3$. It is possible that tail associations in negatively 
related vital rates produce different outcomes for extinction risk
from what we have observed by studying positively associated vital rates.

<!--Tuljapurkar approximation-->
A second topic of research that our work suggests has to do with Tuljapurkar's
approximation of the long-term stochastic growth rate.
Our results suggest that tail associations and asymmetries of tail associations
may have limited influence on the extinction risk of 
stage-structured, non-spatial models. Although we believe more research is 
certainly still needed
as to whether this conclusion holds more generally than the model we considered, 
if it does hold 
generally, it may be possible to obtain a better understanding of 
the result using the well-known approximation
of Tuljapurkar [@tulja1982; @tulja1990] for the long-term stochastic growth rate.
The long-term stochastic growth rate is an important quantity in demography
[@Caswell2000] and population viability analysis for conservation 
[@MorrisDoak2002], and it factors prominently into
extintion risk estimates for density-independent age- or 
stage-structured population models 
[@Lande1988], $\vec{P}(t+1)=M(t)\vec{P}(t)$. Here $\vec{P}(t)$ is the 
age- or stage-structured population and $M$ is a matrix-valued stationary stochastic 
process. The matrix entries of $M$ can correspond, for instance, to the fluctuating
fecundity and survival rates of the life stages of the population. 
For such a model, Tuljapurkar's approximation expresses the long-term stochastic 
growth rate as a function of the variances and covariances of the entries
of $M$ and of the eigenvalues and eigenvectors of the expected value of 
$M$. None of these quantities depends on tail associations between matrix entries.
Thus, whenever Tuljapurkar's approximation is a good
approximation, the long-term stochastic growth rate should not depend sensitively
on tail association. @Lande2003 provides reasoning supporting a claim that 
Tuljapurkar's approximation
is "likely to be accurate in many cases," (p. 60) though we are unaware of the extent
to which accuracy of the approximation has been tested for scenarios with
asymmetric tail associations. It seems worth explicitly exploring the accuracy of 
Tuljapurkar's approximation under scenarios of tail association in the vital rates
and relating results to the effects of tail associations on the long-term
stochastic growth rate and extinction risks.
We note that the above reasoning pertains only to density-independent models,
and says nothing about our results for the density-dependent version of our
stage-structured model.

<!--relate things to the geography of synchrony-->
Finally, our results suggest that combining
ideas about tail association with recently developed ideas about the geography 
of synchrony may provide fertile territory for future research. 
Our metapopulation models assumed that habitat patches were equispaced on a transect or
were all equally mutually accessible through dispersal.
Furthermore, associations between environmental variables in different
patches were assumed to be spatially homogeneous. In real metapopulations,
neither of these assumptions holds. Several recent studies, reviewed and synthesized
by @walteretal2017, have explored geographic variation in the synchrony of environmental
and population variables; @walteretal2017 explored the mechanisms, consequences
and ecological importance of the geography of synchrony. For instance, synchrony was found to 
have a prominent geography which can help provide inferences about the causes
of synchrony and about organism ecology. Two among many research questions that 
could be explored that blend the geography of synchrony with tail associations
include: 1) what is the nature of the geography of tail associations of 
environmental and ecological variables? and 2) do asymmetries of tail
associations of environmental variables interact with geographies of synchrony
to influence metapopulation extinction risk?
Connecting the ideas of tail association that
we have developed with the ideas on the geography of synchrony
of @walteretal2017 seems like a promising future direction.

@@

### 2019-11-08 SuppMat.Rmd
adding new csl
finished incorporating Lawrence comments

---
title: "Supporting information: Tail associations and their impact on extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: ""
geometry: "left=1in,right=1in,top=1in,bottom=1in"

@@

<!--FIG: Verhulst model, local dispersal-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=12 cm]{./Results/common_legend_extreme_and_moderate_noise.pdf}\\
\includegraphics[width=12 cm]{./Results/verhulst_scl_0.6/verhulst_ext_risk_local_disp_extreme_and_moderate_noise.pdf}\\
\caption[Extinction risk v. tail association, $d$ and $r$: metapopulations, local dispersal, Velhurst]{Analogous to Fig. \ref{MT-fig_extrisk_allmodel_localdisp} in the main text, but considering more values of $r$ and only the Velhurst model. \label{fig_risk_local_disp_verhulst_r_vary_scl_0.6}}
\end{center}
\end{figure}

@@

<!--FIG: Verhulst model, global dispersal-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=12 cm]{./Results/common_legend_extreme_and_moderate_noise.pdf}\\
\includegraphics[width=15 cm]{./Results/verhulst_scl_0.6/verhulst_ext_risk_global_disp_extreme_and_moderate_noise.pdf}\\
\caption[Extinction risk v. tail association, $d$ and $r$: metapopulations, global dispersal, Velhurst]{Analogous to Fig. \ref{fig_risk_local_disp_verhulst_r_vary_scl_0.6} but for global dispersal between habitat patches. \label{fig_risk_global_disp_verhulst_r_vary_scl_0.6}}
\end{center}
\end{figure}

@@

\rho=\frac{\sum_{i=1}^N (x_i - \mean(x_i))(y_i-\mean(y_i))}{(N-1)\sqrt{\var(x_i)\var(y_i)}}.

@@

\noindent However, it is straightforward to see that $\mean(x_i) = \mean(y_i)=(N+1)/2$,

@@

$\cov_{i=N/2+1}^N (x_i,y_i)$. To see this, note that

@@

**The Verhulst model.** The Velhurst model 
[@cohen1995unexpected; @may1976bifurcations] is a discrete-time 
representation of logistic growth, and can be written 
as $P(t+1) = F(P(t),r,K) =P(t)[1+r(1-P(t)/K)]$, with $r>0$ and $K>0$ being growth rate and 
carrying capacity parameters, respectively. 
The model equilibrium $P_e=0$ is always unstable whereas the equilibrium 
$P_e=K$ is stable below the bifurcation 
point $r_{bf}=2$. The model shows undercompensatory dynamics for $0<r<r_c=1$
and overcompensatory dynamics for $r_c<r<r_{bf}$.

>>>>

---
title: "Supporting information: Tail associations and their impact on extinction risk"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Daniel C. Reuman"
date: ""
geometry: "left=1in,right=1in,top=1in,bottom=1in"

@@

<!--FIG: Verhulst model, local dispersal-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=12 cm]{./Results/common_legend_extreme_and_moderate_noise.pdf}\\
\includegraphics[width=12 cm]{./Results/verhulst_scl_0.6/verhulst_ext_risk_local_disp_extreme_and_moderate_noise.pdf}\\
\caption[Extinction risk v. tail association, $d$ and $r$: metapopulations, local dispersal, Verhulst]{Analogous to Fig. \ref{MT-fig_extrisk_allmodel_localdisp} in the main text, but considering more values of $r$ and only the Verhulst model. \label{fig_risk_local_disp_verhulst_r_vary_scl_0.6}}
\end{center}
\end{figure}

@@

<!--FIG: Verhulst model, global dispersal-->
\begin{figure}[!h]
\begin{center}
\includegraphics[width=12 cm]{./Results/common_legend_extreme_and_moderate_noise.pdf}\\
\includegraphics[width=15 cm]{./Results/verhulst_scl_0.6/verhulst_ext_risk_global_disp_extreme_and_moderate_noise.pdf}\\
\caption[Extinction risk v. tail association, $d$ and $r$: metapopulations, global dispersal, Verhulst]{Analogous to Fig. \ref{fig_risk_local_disp_verhulst_r_vary_scl_0.6} but for global dispersal between habitat patches. \label{fig_risk_global_disp_verhulst_r_vary_scl_0.6}}
\end{center}
\end{figure}

@@

\rho=\frac{\sum_{i=1}^N (x_i - \mean(x_i))(y_i-\mean(y_i))}{(N-1)\sqrt{\var(x_i)\var(y_i)}},

@@

\noindent where $\var$, here and throughout, is the sample variance, with Bessel's correction 
(denominator $N-1$). However, it is straightforward to see that $\mean(x_i) = \mean(y_i)=(N+1)/2$,

@@

$\cov_{i=N/2+1}^N (x_i,y_i)$ (again using $N-1$ in the denominator, here and throughout). 
To see this, note that

@@

**The Verhulst model.** The Verhulst model 
[@cohen1995unexpected; @may1976bifurcations] is a discrete-time 
representation of logistic growth, and can be written 
as $P(t+1) = F(P(t),r,K) =P(t)[1+r(1-P(t)/K)]$, with $r>0$ and $K>0$ being growth rate and 
carrying capacity parameters, respectively. 
The model equilibrium $P_e=0$ is always unstable whereas the equilibrium 
$P_e=K$ is stable below the bifurcation 
point $r_{bf}=2$. The model shows undercompensatory dynamics for $0<r<r_c=1$
and overcompensatory dynamics for $r_c<r<r_{bf}$.
